\documentclass[11pt, a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath, amssymb}
\usepackage{hyperref}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{geometry}
\usepackage{booktabs}
\usepackage{float}
\usepackage{enumitem}
\usepackage{setspace}
\usepackage[style=apa, backend=biber, eprint=false]{biblatex}

% Formatting Requirements
\geometry{margin=1in, a4paper}
\onehalfspacing % 1.5 line spacing

\addbibresource{../references.bib}

\begin{document}

% Title Page
\begin{titlepage}
    \centering
    \vspace*{1cm}

    \Large
    \textsc{Thesis Proposal}

    \vspace{0.5cm}
    
    \Huge
    \textbf{Evolving Deception: Efficacy of Evolutionary Algorithms in Automated Red-Teaming for Instrumental Deception}
    
    \vspace{1.5cm}
    
    \Large
    \textbf{Elias Schlie} \\
    Student Number: u783766 \\
    \texttt{e.p.schlie@tilburguniversity.edu}
    
    \vspace{2cm}
    
    Supervisor: Dr. Emmanuel Keuleers \\
    
    \vfill
    
    \large
    Tilburg University \\
    School of Humanities and Digital Sciences \\
    Department of Cognitive Science and Artificial Intelligence \\
    
    \vspace{2cm}

    Tilburg, The Netherlands \\
    \today
    
\end{titlepage}

\newpage

\section{Project Definition}
% Task: Provide a clear description of the problem you plan to address.

As large language models (LLMs) are put into increasingly critical positions in society, we need scalable methods to track whether and in which situations they lie to us.
Current approaches to find such scenarios often lean on zero or multi-shot generation.
While showing some success, those methods often require heavy human intervention to ensure breadth and realism of generated scenarios \parencite{renMASKBenchmarkDisentangling2026}. 
A potential pathway to address this is to evolutionarily inject previous successsful examples into the prompt of the generator.
This thesis will test the viability of such an evolutionary approach. 
Specifically, we are going to address the research question: 

\begin{quote}
    \itshape To what extent does evolutionary optimization outperform zero and multi-shot generation in discovering realistic scenarios that induce instrumental deception in safety-aligned LLMs?
\end{quote}

By comparing how success-rate, scenario realism, and scenario diversity change when introducing evolution, we are trying to get an accurate picture whether evolution is a valuable addition for this problem.

\section{Motivation}
% Task: Briefly explain why this problem is worth addressing from both societal and scientific perspectives.

Researching efficient ways to audit LLMs for deceptive behaviors is becoming increasingly relevant from both a societal and scientific perspective.

Societally, there is a critical need for users to know how much they can trust the systems they are using.
This trust is not possible if models deceive and lie to users.
Without proper auditing, it is possible that we are not aware of deceptive tendencies in models and start using such a model in contexts where deceptive behavior can cause critical damage.
Currently, such audits are mostly bottlenecked by time of human experts.
Testing whether evolutionary methods increase efficacy of those audits could uncover a way to decrease the amount of human intervention necessary for each alignment audit.
This productivity enhancement could then be invested into doing more or creating better deception audits.
Users would gain more access to auditing scores, and could use them to choose the best model, decreasing risks.
This would have the additional benefit of incentivizing model providers to invest more heavily in improving the honesty of their models.

Scientifically, this research aims to validate evolutionary algorithms as a superior method for red-teaming.
If successful, this approach enables the autonomous creation of robust honesty benchmarks.
Such benchmarks facilitate the rigorous evaluation of interventions designed to reduce deceptive behavior.
Crucially, as models improve and static benchmarks lose signal, evolutionary methods can dynamically generate harder scenarios to maintain evaluative pressure.
The adversarial nature of this approach also helps discover novel failure modes arising from specific training methods or scaling.
By identifying these vulnerabilities, researchers can make informed decisions about model deployment and refinement.
Finally, the generated data supports the adversarial training of honesty classifiers, such as probes, enhancing their accuracy and robustness.

\section{Background}
% Task: Summarize what is known in the scientific literature about this problem. 
% Base this on at least 5 relevant sources (last 5 years, peer-reviewed).

This thesis combines two fields in the LLM literature. First, it addresses deception \parencite{parkAIDeceptionSurvey2024}, which has been explored and measured from many different angles \parencite{renMASKBenchmarkDisentangling2026,huangDeceptionBenchComprehensiveBenchmark2025, wangEvaluatingHonestyLie2025}. 
Secondly, it leans heavily on insights from automated jailbreaking \parencite{chenJailbreakingLLMsVLMs2026}, which uses adversarial algorithms to find situations in which LLMs comply with harmful user requests and alignment auditing \parencite{marksAuditingLanguageModels2025}, which tries to detect hidden misaliged goals and behaviors in LLMs.

\subsection{Deception}

Following the \textcite{renMASKBenchmarkDisentangling2026}, we define instrumental deception as goal-directed, strategic lies of commission, which are untrue statemets, made with the intent to deceive \parencite{mahonDefinitionLyingDeception2016}.
This differs from hallucinations \parencite{rawteTroublingEmergenceHallucination2023}, where the model lacks correct knowledge and from sycophancy \parencite{sharmaUnderstandingSycophancyLanguage2023}, where the model passively echoes user views rather than actively trying to steere them.

Several methods attempt to measure lying and deception in LLMs.
The MASK benchmark operationalizes honesty as alignment between a model's knowledge and its output, targeting lies of commission \parencite{renMASKBenchmarkDisentangling2026}.
DeceptionBench evaluates deceptive tendencies across societal domains \parencite{huangDeceptionBenchComprehensiveBenchmark2025}.
\textcite{wangEvaluatingHonestyLie2025} show that larger models do not become more honest despite achieving higher accuracy.
Each of those papers relies on a mix of LLM generation and human review to create their test cases.
To create the MASK benchmark, for instance, authors hired and trained a group of contributors, which manually generated and checked examples at different states of the process. 

A big concern when testing for deception in LLMs, is evaluation awareness \parencite{needhamLargeLanguageModels2025}.
Evaluation awareness describes the ability of LLMs to recognize when they are being tested and potentially changing their behavior based on this.
Through a process known as alignment faking, LLMs strategically modify their outputs to appear compliant with training objectives during assessment.
However, these models revert to misaligned behavior once they detect that oversight is absent or unlikely \parencite{greenblattAlignmentFakingLarge2024}.
Behavioral observations from unrealistic scenarios might thus not be trustable. 
To address this, we measure scenario realism in our evaluation and explicitly include it in the fitness function of the evolutionary approach.

\subsection{Red-teaming and Alignment Auditing}

Manual red-teaming is expensive and bottlenecked by human time \parencite{hongCuriositydrivenRedteamingLarge2023}.
Automated approaches use reinforcement learning, iterative optimization, or evolutionary algorithms to generate adversarial prompts more efficiently \parencite{perezRedTeamingLanguage2022,liuAutoDANTurboLifelongAgent2025,samvelyanRainbowTeamingOpenEnded2024}.
Evolutionary methods are particularly appealing because they maintain diverse populations of candidate attacks and iteratively refine them according to domain-specific fitness functions \parencite{yuLLMVirusEvolutionaryJailbreak2024,chengForgeDANEvolutionaryFramework2025}.
However, existing pipelines optimize for generic policy violations rather than for deception specifically \parencite{jauhariAlgorithmicRedTeaming2025}.
This thesis addresses that gap by targeting instrumental deception as the optimization objective and encoding realism into the fitness function.

Finally, the proposed methodology uses LLMs not only as targets but also as generators and judges.
While LLM judges can approach human-level agreement on well-defined tasks, their reliability degrades for nuanced or contextual criteria \parencite{jauhariAlgorithmicRedTeaming2025}.
This motivates the cross-model validation and human-annotated subset described in Section~5.

\section{Experimental Design}
% Task: Describe the design of your experiment or the dataset(s) you will use in as much detail as possible (size, format, accessibility). 
% Provide a rationale for choosing these data.

This study employs a comparative experimental design to evaluate the efficacy of evolutionary algorithms.
We compare zero and multi-shot baselines against an evolutionary optimization approach.

A judge LLM evaluates each scenario on two dimensions.
First, Deception Success is a binary metric: 1 if the target commits a lie of commission, 0 otherwise.
Second, Realism is assessed on a 1-7 Likert scale based on detailed criteria.
We define the composite fitness metric as the product of these scores ($Fitness = Realism \times DeceptionSuccess$).
This multiplicative approach ensures that scenarios failing to elicit deception receive a fitness of zero, regardless of their realism.

The experiment compares three conditions:
\begin{enumerate}
    \item \textbf{Zero-shot}: 
        The generator LLM produces deceptive scenarios directly based on a prompt, explaining the task. 
        This serves as the unoptimized baseline.
    \item \textbf{Multi-shot}: 
        Adds a static set of manually curated examples to the generation prompt. 
        This evaluates the impact of simple in-context learning.
    \item \textbf{Evolutionary}: 
        After initializing a population of generated scenarios in a zero-shot fashion, we add a subset, selected from this population, as examples to the generation prompt. 
        This subset is randomly selected for every new generation attempt with the probability of any example getting selected being proportional to that example's fitness.
\end{enumerate}

Each condition will run for an equivalent amount of time to ensure a fair comparison.

\section{Statistics, Algorithms, and Software}
% Task: Describe the statistical analyses or data mining and machine learning algorithms and software you plan to use. 
% Explain evaluation metrics, baseline methods, and ground-truth labeling plans.

We will use GLM-4.7-Flash for all three roles: Generator, Target, and Judge.
This unified setup maximizes computational efficiency.
However, it risks self-preference bias and overfitting.
To control for this, we perform two validation checks.
First, we run a cross-model validation using Qwen3-VL-32B.
Qwen3 acts as both Target and Judge on a subset of final scenarios to test generalizability.
Second, we validate the Judge's reliability against human ground truth.
We manually annotate a random sample of 50 scenarios for realism and deception success.
We calculate the alignment between human and automated judgements to ensure metric validity.
For efficient inference, we use the vLLM library.

We will plot average deception elicitation success, realism of the top 100 generated scenarios over time for each method.
To answer the research question, we compare the final populations of each condition.
We analyze Deception Success Rate (binary) using the Chi-Square test for independence, with pairwise comparisons via Fisher's Exact Test (Bonferroni-corrected).
For Realism (ordinal), we use the Kruskal-Wallis H test to detect overall differences, followed by post-hoc Mann-Whitney U tests with Bonferroni correction.
All statistical analyses are performed using Python libraries such as SciPy and Scikit-learn.

\section{Milestones and Plan}
% Task: Outline the major intermediate milestones and provide a general idea of your planning.

\begin{table}[H]
    \centering
    \caption{Project Milestones and Timeline}
    \begin{tabular}{@{}ll@{}}
        \toprule
        Milestone & Timeline \\
        \midrule
        Literature Review & Week 1-2 \\
        Framework implementation & Week 3-4 \\
        Running experiments & Week 5 \\
        Quantitative Evaluation \& Analysis & Week 6 \\
        Thesis Writing \& Finalization & Week 7-9 \\
        \bottomrule
    \end{tabular}
\end{table}

\section*{Acknowledgements}
Claude Opus 4.6 \parencite{anthropicClaudeOpus462026} was used for language rephrasing; all content was human-reviewed and edited for accuracy.

\newpage
\printbibliography

\end{document}
